---
title: "NeedlemanWunsch"
author: "Beatrice Baldassarre"
date: "`r Sys.Date()`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{NeedlemanWunsch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup}
library(NeedlemanWunsch)
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```



# Introduction

Sequence alignment is a way of arranging sequences of DNA, RNA, or protein to identify regions of similarity that may be a consequence of functional, structural, or evolutionary relationships between the sequences. The alignment of exactly two sequences is called pairwise alignment, and it can be performed globally or locally.  The Needleman-Wunsch package provides functions to perform a pairwise global sequence alignment using the Needleman-Wunsch algorithm.

# Functions

## Align two sequences

Aligning two sequences using the *Needleman-Wunsch algorithm* is now an easy task. With the **globalAlignnmentNeedlemanWunsch** function. What the user needs are the costs of matches and mismatches, the gap penalty and the two sequences.

```{r}
# Perform global sequence alignment
globalAlignnmentNeedlemanWunsch <- function(matchCost, mismatchCost, gapPenalty, sequenceA, sequenceB) {
    scoresMatrix <- creatingMatrix(matchCost, mismatchCost, gapPenalty, sequenceA, sequenceB)
    alignment <- performTraceback(scoresMatrix,sequenceA,sequenceB)

    result <- list(scoresMatrix = scoresMatrix,
                   matrixDirection = alignment$matrixDirection,
                   alignmentA = alignment$alignmentA,
                   alignmentB = alignment$alignmentB)
    return(result)
}
print(globalAlignnmentNeedlemanWunsch(7,-3,-4,"GTT","GCATT"))
```

## Creating the scores matrix

The first step of the Needleman-Wunsch algorithm is the creation of *scores matrix*. The function **creatingMatrix** computes this matrix. As for the globalAlignmentNeedlemanWunsch function, what the user needs are the costs of matches and mismatches, the gap penalty and the two sequences.

```{r}
# Creation of the scores matrix
creatingMatrix <- function(matchCost, mismatchCost, gapPenalty, sequenceA, sequenceB) {
    matrix <- matrix(0, nrow = nchar(sequenceA)+1, ncol = nchar(sequenceB)+1)

    for(i in 2:nrow(matrix)) {
        matrix[i,1] <- gapPenalty * (i-1)
    }

    for(j in 2:ncol(matrix)) {
        matrix[1,j] <- gapPenalty * (j-1)
    }

    for(i in 2:nrow(matrix)) {
        for(j in 2:ncol(matrix)){
            top <- matrix[i-1,j] + gapPenalty
            left <- matrix[i,j-1] + gapPenalty
            if (isItAMatch(sequenceA,sequenceB,i,j))    diagonal <- matrix[i-1,j-1] + matchCost
            else    diagonal <- matrix[i-1,j-1] + mismatchCost
            matrix[i,j] <- max(top,left,diagonal)
        }
    }
    return(matrix)
}
print(creatingMatrix(7,-3,-4,"GTT","GCATT"))

```

## Traceback 

In order to align two sequences, the creation of the scores matrix is followed by the *traceback* step. The function **performTraceback** generates a matrix with the directions of the alignment while computing the alignment itself. It takes as input a scores matrix and two sequences. 
```{r}
# Traceback step
performTraceback <- function(matrix, sequenceA, sequenceB) {
    matrixDirection <- matrix
    alignmentA <- ''
    alignmentB <- ''
    i <- nrow(matrixDirection)
    j <- ncol(matrixDirection)
    while (i>1 | j>1) {
        matrixDirection[i,j] <- "*"
        if(isItAMatch(sequenceA,sequenceB,i,j)) {
            alignmentA <- paste(substring(sequenceA, i-1, i-1),alignmentA)
            alignmentB <- paste(substring(sequenceB, j-1, j-1),alignmentB)
            i <- i-1
            j <- j-1
        }
        else {
            top <- matrix[i-1,j]
            left <- matrix[i,j-1]
            diagonal <- matrix[i-1,j-1]

            if (top == max(c(top,left,diagonal))) {
                alignmentA <- paste(substring(sequenceA, i-1, i-1),alignmentA)
                alignmentB <- paste("-",alignmentB)
                i <- i-1
            }

            else if (left == max(c(top,left,diagonal))){
                alignmentB <- paste(substring(sequenceB, j-1, j-1),alignmentB)
                alignmentA <- paste("-", alignmentA)
                j <- j-1
            }

            else  {
                alignmentA <- paste(substring(sequenceA, i-1, i-1),alignmentA)
                alignmentB <- paste(substring(sequenceB, j-1, j-1),alignmentB)
                i <- i-1
                j <- j-1
            }

        }

    }
    result <- list(matrixDirection = matrixDirection,
                   alignmentA = alignmentA,
                   alignmentB = alignmentB)
    return(result)
}
```
